Gradual Soundness: Lessons from Static Python

### 2021-12-31 pj-7.1/paper.tex

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\defeq}{=}
\newcommand{\mfeq}{=}
\newcommand{\langmid}{\mathrel{\mathbf{\Big\vert}}}
\newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}

@@

\subsection{Dynamic Python Features}

@@

@@

\subsection{Analyzing Code Changes}

The fact that the \SP{} team has been changing code as well as types
during its migrations is potentially troubling.
It could be that the \SP{} approach to gradual typing only pays off
after significant modifications, or only for a restrictive style of programming.

To understand the code changes, an author affiliated with Instagram reviewed
approximately 30 patches that significantly improved performance.
In general, the patches contain relatively minor code changes.
The biggest changes are to testing code rather than production code.
Both these points are encouranging.
The following nine points describe common patterns:

\begin{enumerate}
  \item
   adjustments to tests to pass correct types instead of mock objects, as as
   not to fail type checks in SP code. This is an inherent and desired
   incompatibility; next half we plan to work on improving the Python mock
   framework to automatically create mocks of the correct types.
   %% - bg: lookup the dino email about mocks

 \item
   adjustments to mock assertions in tests about calls to adjust for the
   fact that SP currently converts all call arguments to be positional (we plan
   to fix this incompatibility by falling back to the original call arguments
   when we detect the target function has been patched.)
   %% - bg: ??? bit fuzzy about positional and others

 \item
   adjustments around class attributes vs instance attributes (Python allows
   this to be a bit fuzzy, where a class attribute can be a fallback default
   for an instance attribute if not set, but SP currently requires an attribute
   to be clearly one or the other, and we have more efficient access for
    instance attributes via \code{LOAD\_FIELD} that we don’t have for class attributes.)
   %% - related to github #37 ?

 \item
   convert @classmethod that don’t actually need the class to @staticmethod,
    allowing for emitting direct \code{INVOKE\_FUNCTION} instead of \code{INVOKE\_METHOD}

 \item
   extract special types that our compiler doesn’t yet support/understand
   (classes decorated with @dataclass or inheriting from `enum.Enum`) out of
   the module we are converting to static. We have work in progress on
   supporting some Enum types, and plan to add a @dataclass intrinsic soon.

 \item
   Removing the use of super-dynamic Python features like \code{\_\_setattr\_\_} that
   we don’t support. So far in the cases where we’ve done this it hasn’t
   required a major rewrite.
   %% - bg: more details pls ,,, sketch the rewrite

 \item
   Removing the use of keyword-only arguments (bare `*`) in function
   signatures, since we don’t yet support calling such functions. We plan to
   support this next half.

 \item
   Convert Python integers and bools used in very hot code paths into SP
   machine primitives.
   %% - bg: focus on block, convert, no problem?

 \item
   Occasionally converting a regular Python list or dict to a CheckedList or
   CheckedDict.
   %% - bg: how occasional? worried about power of the "gradual soundness" claim
\end{enumerate}

%% significant change: replaced `contextlib.ContextDecorator` with `__static__.ContextDecorator`
%%  removes an extra call layer
%%  helps with (timing [ context managers / decorators ])
%%
%% all of this is the low-hanging fruit, haven't tackled uses of metaclasses even though looks promising
%%  because the pay / payoff isn't yet attractive

\subsection{Developer Experience}

%% - pushback one time about removing MI multiple inheritance
%% - otherwise, people happy with low-cost improvements

%% - pyre challenges?

FILL

@@

\begin{table}[t]
  \caption{Microbenchmark Performance Ratios}
  \label{t:microbenchmark}
  %% show untyped Python time?
  \begin{tabular}{ll}
    Name               & Typed / Dynamic \\\midrule
    \bmname{deltablue} &             0\% \\
    \bmname{fannkuch}  &             0\% \\
    \bmname{nbody}     &             0\% \\
    \bmname{nqueens}   &             0\% \\
    \bmname{pystone}   &             0\% \\
    \bmname{richards}  &         0.125\% \\
  \end{tabular}
\end{table}

@@

%% Immediate future: fix adoption friction points
%% 2021-12-15: WHAT ARE THESE?

>>>>

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\defeq}{=}
\newcommand{\mfeq}{=}
\newcommand{\langmid}{\mathrel{\mathbf{\Big\vert}}}
\newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
\newcommand{\ipara}[1]{\emph{#1}\/.}

@@

\subsection{Dynamic Python Features}
\label{s:dynamic-python}

@@

\paragraph{Set attribute}

Carl: Removing the use of super-dynamic Python features like
\code{\_\_setattr\_\_} that we don’t support. So far in the cases where
we’ve done this it hasn’t required a major rewrite.

@@

\subsection{Analyzing Code Changes}
%% TODO boring unclear writing

Because the \SP{} team has been changing code as well as types
during its migrations, the question arises as to whether the code changes
raise any threats to validity.
It could be that the \SP{} approach to gradual typing only pays off
after significant modifications, or only for a restrictive style of programming.

To understand the code changes, we reviewed approximately 30 patches that
significantly improved performance.
In general, the patches have minor code changes and most of these changes
affect tests rather than production code.
Digging deeper, there are nine common kinds of changes.
Six of these address type errors; the other three focus on performance.

%% one significant change: replaced `contextlib.ContextDecorator` with `__static__.ContextDecorator`
%%  removes an extra call layer
%%  helps with (timing [ context managers / decorators ])

%% TODO section formatting looks good?
\subsubsection{Code Changes for the Type Checker}

    %% X bg: lookup the dino email about mocks
    %% And we do go through great lengths to make monkey patching work.  Because
    %% we need to enforce return types it ends up being not quite as flexible as
    %% normal Python.  For example one common pattern is to use Python’s mock
    %% library and we currently end up throwing lots of TypeError’s when a
    %% function which returns None is mocked and starts returning a MagicMock
    %% object instead.  But it’s an easy fix in that you can specify the return
    %% value to be compatible.  There may be more complex cases where you want to
    %% return a mock for an actual class.  For example in our code base there’s
    %% frequent mocking of the HttpRequest object and we have a MockReqeust object
    %% which is a subclass of HttpRequest so that these can still be mocked
    %% (luckily this actually pre-dated our static conversion work).

\begin{enumerate}
  \item
    \ipara{Fix type errors due to mock wrappers}
    When test code introduces a mock wrapper, it may change the return type of
    a function.
    For example, the wrapped version of a function that returns \code{None}
    will return a \code{MagicMock} object unless the programmer explicitly
    specifies a return value.
    %% Carl: This is an inherent and desired
    %% incompatibility; next half we plan to work on improving the Python mock
    %% framework to automatically create mocks of the correct types.

  \item
    \ipara{Change mocked function to expect positional arguments}
    \SP{} currently rewrites all function calls to use positional arguments.
    Mock-wrapped functions therefore need to positional arguments instead of keywords.
    %% Carl: we plan to fix this incompatibility by falling back to the original call
    %% arguments when we detect the target function has been patched.

  \item
    \ipara{Change functions to accept positional arguments}
    Functions that expect only keyword arguments cannot yet be called from \SP{} contexts.
    These functions need to change.
    %% Carl: We plan to support this next half.

  \item
    \ipara{Organize class and instance attributes}
    Whereas Python allows class attributes to serve as default values for instance attributes,
    \SP{} does not.
    Programmers have to decide where an attribute belongs so that \SP{} can optimize
    reads from instance attributes.
    %% Carl: (efficient via \code{LOAD\_FIELD})
    %% - related to github #37 ?

  \item
    \ipara{Move \code{@dataclass} and \code{Enum} classes}
    \SP{} does not support classes decorated with \code{@dataclass}
    and classes that are enums.
    Programmers have to move these classes to a different module for now.
    %% Carl: We have work in progress on supporting some Enum types, and plan to add a @dataclass intrinsic soon.

  \item
    \ipara{Remove unsupported Python features}
    Occurrences of the dynamic Python features listed in \cref{s:dynamic-python} must be rewritten or moved.
    According to the \SP{} team, these changes have not required any major rewrites.

\end{enumerate}

%% TODO section formatting looks good?
\subsubsection{Code Changes for Performance}

\begin{enumerate}
  \item
    \ipara{Convert \code{@classmethod} to \code{@staticmethod}}
    Static methods can get invoked directly as functions, bypassing the class.
    %% Carl: \code{INVOKE\_FUNCTION} instead of \code{INVOKE\_METHOD}

  \item
    \ipara{Use primitive types}
    Any Python integers and booleans in hot code paths are better as
    \SP{} primitive values~(\cref{s:c-types}).
    %% - bg: focus on block, convert, no problem?

  \item
    \ipara{Use Checked Types}
    Occasionally, converting a Python dictionary or Python list to a
    checked type has a big impact on performance.
    As noted in \cref{s:checked-type}, these conversions affect all downstream
    clients of the data structure.
    %% - bg: how occasional? worried about power of the "gradual soundness" claim

\end{enumerate}

\subsubsection{Developer Experience}

For the most part, developers at Instagram have been happy adopting \SP{}.
The transition from Pyre to \SP{} has thus far delivered performance benefits
that outweigh the cost of code changes.

One significant pushback dealt with multiple inheritance.
The \SP{} team proposed a rewrite from multiple inheritance to single inheritance
to improve performance and static checks.
But the code maintainers preferred to keep multiple inheritance.
Because \SP{} defaults to Python behavior, the maintainers were able
to keep multiple inheritance alongside \SP{} code.

%% any more?

@@

\begin{table}[t]
  \caption{Microbenchmark Performance Ratios}
  \label{t:microbenchmark}
  %% show untyped Python time?
  %% transpose?
  \begin{tabular}{ll}
    Name               & Typed / Dynamic \\\midrule
    \bmname{deltablue} &             0\% \\
    \bmname{fannkuch}  &             0\% \\
    \bmname{nbody}     &             0\% \\
    \bmname{nqueens}   &             0\% \\
    \bmname{pystone}   &             0\% \\
    \bmname{richards}  &         0.125\% \\
  \end{tabular}
\end{table}

@@

%% Immediate future: fix adoption friction points
%% 2021-12-15: WHAT ARE THESE?
%% - mocks need manual return_value ... "next half we plan to work on improving the Python mock framework to automatically create mocks of the correct types."
%% - mock calls have to rewrite to expect positional args, not keywords "plan to avoid rewrite if fn is patched"
%% - functions have to expect positionals (can't do keyword only)
%% - @dataclass enum.Enum
%%
%% significant selection bias
%% all of this is the low-hanging fruit, haven't tackled uses of metaclasses even though looks promising
%%  because the pay / payoff isn't yet attractive
